---
title: "FlightDelay"
author: "Group 6"
date: "2022-12-07"
output: pdf_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(RSQLite)
library(modelr)
library(arrow)
library(dplyr)
```
Reading in flight data from 2018 - 2022 using smaller .parquet files to limit size
```{r}
#Data from Kaggle: https://www.kaggle.com/datasets/robikscube/flight-delay-dataset-20182022?select=Combined_Flights_2022.csv
con <- dbConnect(SQLite(), "flight_data.db")
```

```{r}
path <- "C:/Users/chase/OneDrive/Northeastern Code/DS 5110/Group Project/data/Combined_Flights_2021.parquet"
flights_2021 <- read_parquet(path, as_data_frame = TRUE) #read in flight data
dbWriteTable(con, "flights", flights_2021) #initializing flights df
rm(flights_2021) #dropping df from environment to minimize memory usage
```

```{r} 
df <- as_tibble(dbGetQuery(con, 
           "SELECT FlightDate
                 , Airline
                 , Origin
                 , Dest
                 , Cancelled
                 , Diverted
                 , DepTime
                 , DepDelayMinutes
                 , Distance
                 , DistanceGroup
                 , Year
                 , Quarter
                 , Month
                 , DayofMonth
                 , DayofWeek
                 , Marketing_Airline_Network
                 , OriginAirportID
                 , OriginCityName
                 , OriginStateName
                 , OriginWAC
                 , DestAirportID
                 , DestCityName
                 , DestStateName
                 , DestWac
                 , CRSDepTime
                 , CRSElapsedTime
                 , ArrDelayMinutes
                 FROM flights
           WHERE Operating_airline in ('AA', 'UA', 'DL')
          AND OriginWac BETWEEN 1 AND 93
          AND DestWac BETWEEN 1 AND 93"))
df$OriginEncode <- as.numeric(as.factor(df$Origin))
df$DestEncoder <- as.numeric(as.factor(df$Dest))
ggplot(df, aes(x=DepDelayMinutes)) +
  geom_histogram() + 
  theme_minimal()
df <- na.omit(df)
```
We can see that departure delay is very frequently zero or near zero
```{r}
#Compartmentalize States into Census defined regions for analysis
Northeast <- c("Connecticut", "Maine", "Massachusetts", "New Hampshire", "Rhode Island", "Vermont", "New Jersey", "New York", "Pennsylvania", "Washington D.C.")
Midwest <- c("Illinois", "Indiana", "Michigan", "Ohio", "Wisconsin", "Iowa", "Kansas", "Minnesota", "Missouri", "Nebraska", "North Dakota", "South Dakota")
South <- c("Florida", "Georgia", "North Carolina", "South Carolina", "Virginia", "West Virginia", "Alabama", "Kentucky", "Mississippi", "Tennessee", "Arkansas", "Louisiana", "Oklahoma", "Texas", "Delaware", "Maryland", "U.S. Virgin Islands", "Puerto Rico")
West <- c("Arizona", "Colorado", "Idaho", "Montana", "Nevada", "New Mexico", "Utah", "Wyoming", "Alaska", "California", "Hawaii", "Oregon", "Washington", "U.S. Pacific Trust Territories and Possessions")
df <- df %>% 
  mutate(OriginRegion = case_when(
    OriginStateName %in% Northeast ~ "Northeast",
    OriginStateName %in% Midwest ~ "Midwest",
    OriginStateName %in% South ~ "South",
    OriginStateName %in% West ~ "West"
  ), .after = OriginStateName
)
df <- df %>%
  mutate(DestRegion = case_when(
    DestStateName %in% Northeast ~ "Northeast",
    DestStateName %in% Midwest ~ "Midwest",
    DestStateName %in% South ~ "South",
    DestStateName %in% West ~ "West"
  ), .after = DestStateName
)
#Partitioning data
df_part <- resample_partition(df,
                                    p=c(train=0.5,
                                        valid=0.25,
                                        test=0.25))
```
We re-encode region variables into quadrants based on US Census region data, in order to allow for greater interpretability.
```{r}
# Downsampling with 80% data.
df1 <- (df %>% filter(Airline=="Delta Air Lines Inc.", DepDelayMinutes==0))[427224:534031,]
df2 <- (df %>% filter(Airline=="United Air Lines Inc.", DepDelayMinutes==0))[225760:282200,]
df3 <- (df %>% filter(Airline=="American Airlines Inc.", DepDelayMinutes==0))[377192:471490,]
df4 <-  (df %>% filter(DepDelayMinutes!=0))
df_down <- rbind(df1,df2,df3, df4)
dim(df_down)

interval <- function(x) {
  case_when(
    x == 0 ~ "On Time",
    between(x, 1, 60) ~ "Less Delay",
    between(x, 61, 120) ~ "Medium Delay",
    x >= 121 ~ "Large Delay"
  )
}
df_down$DepDelayclass<-interval(df_down$DepDelayMinutes)
df_down %>%
  select(Airline, DepDelayclass) %>%
  group_by(DepDelayclass, Airline) %>%
  summarise(n())
ggplot(df_down,aes(x = DepDelayclass, fill = Airline)) + geom_bar(stat="count",position = "dodge")
#Partitioning data
df_down_part <- resample_partition(df_down,
                                    p=c(train=0.5,
                                        valid=0.25,
                                        test=0.25))
```
Downsampling here allows us to even out our significant class imbalance
```{r}
#function to perform single step of stepwise model selection using RMSE, inspired by lecture (r markdown) code.
step <- function(response, predictors, candidates, partition)
{
  rhs <- paste0(paste0(predictors, collapse="+"), "+", candidates)
  formulas <- lapply(paste0(response, "~", rhs), as.formula)
  rmses <- sapply(formulas,
                  function(fm) rmse(lm(fm, data=partition$train),
                                    data=partition$valid))
  names(rmses) <- candidates
  attr(rmses, "best") <- rmses[which.min(rmses)]
  rmses
}
```

Stepwise

```{r}
df <- df %>% filter(DepDelayMinutes!=0)
#initalizing model variable
model <- NULL
```
OriginRegion: No Region seems to strongly effect log(DepDelayMinutes)
DestRegion: No Region seems to strongly effect log(DepDelayMinutes)
Airline: Doesn't seem highly significant, though American probably highest
Distance: Hard to tell, but appears fairly linear with log(Distance)
DistanceGroup: Unclear if significant
Quarter: 3 slightly higher, not significant seeming
Month: 6-8 appear slightly higher
DayofMonth: Cant tell if any are higher
DayofWeek: Cant tell if any are higher


Using our forward stepwise function we select the most significant variable for predicting Departure Delay with each iteration
```{r}
preds <- "1"
cands <- c("OriginEncode", "DestEncoder", "Airline", "Distance","DistanceGroup", "Quarter", "Month", "DayofMonth", "DayOfWeek")
s1 <- step("log1p(DepDelayMinutes)", preds, cands, df_down_part)
model <- c(model, attr(s1, "best"))
s1
```


```{r}
preds <- c("Airline")
cands <- c("OriginEncode", "DestEncoder", "Month", "Distance","DistanceGroup", "Quarter", "DayofMonth", "DayOfWeek")
s1 <- step("log1p(DepDelayMinutes)", preds, cands, df_down_part)
model <- c(model, attr(s1, "best"))
s1
```

```{r}
preds <- c("Month", "Airline")
cands <- c("OriginEncode", "DestEncoder", "Distance","DistanceGroup", "Quarter", "DayofMonth", "DayOfWeek")
s1 <- step("log1p(DepDelayMinutes)", preds, cands, df_down_part)
model <- c(model, attr(s1, "best"))
s1
```

```{r}
preds <- c("Month", "Airline", "Quarter")
cands <- c("Distance", "DestEncoder","DistanceGroup", "OriginEncode", "DayofMonth", "DayOfWeek")
s1 <- step("log1p(DepDelayMinutes)", preds, cands, df_down_part)
model <- c(model, attr(s1, "best"))
s1
```

```{r}
preds <- c("Month", "Airline", "Quarter", "OriginEncode")
cands <- c("DestEncoder","DistanceGroup", "Distance", "DayofMonth", "DayOfWeek")
s1 <- step("log1p(DepDelayMinutes)", preds, cands, df_down_part)
model <- c(model, attr(s1, "best"))
s1
```

```{r}
preds <- c("Month", "Airline", "Distance", "OriginEncode", "Quarter")
cands <- c("DestEncoder","DistanceGroup", "DayOfWeek", "DayofMonth")
s1 <- step("log1p(DepDelayMinutes)", preds, cands, df_down_part)
model <- c(model, attr(s1, "best"))
s1
```
```{r}
preds <- c("Month", "Airline", "Distance", "OriginEncode", "Quarter", "DayofMonth")
cands <- c("DestEncoder","DistanceGroup", "DayOfWeek")
s1 <- step("log1p(DepDelayMinutes)", preds, cands, df_down_part)
model <- c(model, attr(s1, "best"))
s1
```
```{r}
preds <- c("Month", "Airline", "Distance", "OriginEncode", "Quarter", "DayofMonth", "DayOfWeek")
cands <- c("DestEncoder","DistanceGroup")
s1 <- step("log1p(DepDelayMinutes)", preds, cands, df_down_part)
model <- c(model, attr(s1, "best"))
s1
```

```{r}
step_model <- tibble(index=seq_along(model),
                     variable=factor(names(model), levels=names(model)),
                     RMSE=model)
ggplot(step_model, aes(y=RMSE)) +
  geom_point(aes(x=variable)) +
  geom_line(aes(x=index)) +
  labs(title="Stepwise model selection") +
  theme_minimal()
```
Here we plot the step-wise function to determine the decrease in RMSE each variable's inclusion gives us.

```{r}
fitting <- lm(log1p(DepDelayMinutes) ~ OriginEncode + Airline + Month + Distance + Quarter, data =df_down_part$train)
rmse(fitting, df_down_part$test)
summary(fitting)
```
We chose to select Origin, Airline, Month, Distance, and Quarter as our variables for prediction, as including the next variables did not give large RMSE decreases, and could result in overfitting.

# Code for DepDelayMinutes=0 condition.
Finally, we consider the effect of removing rows where Departure Delay = 0, this gives us a much more normal distribution of residuals, and shows that it may be good to separate this analysis into two tasks, prediction of Delayed vs Not-Delayed, and separately, if delayed, by how much?
```{r}
# Trying depDelayMinutes=0 condition.
df <- df %>% filter(DepDelayMinutes!=0)
fitting <- lm(log10(DepDelayMinutes) ~ OriginEncode + DestEncoder + Airline + Month + Distance + DayofMonth + DayOfWeek + DepTime, data =df)
df %>% add_residuals(fitting, "resid") %>%
  ggplot(aes(x=resid)) +
  geom_histogram() + 
  theme_minimal()
```


```{r}
df %>%
add_residuals(fitting, "resid") %>%
ggplot(aes(sample=resid)) +
geom_qq() +
labs(title="QQ plot is approximately normal", y="residuals")+
theme(plot.title=element_text(hjust=0.5, color="red"))
```